---
title: "Project 1"
author: "Riddhi"
date: "10/17/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (a) Plot the data and elaborate on the general pattern of the data.

```{r}
setwd("/Users/riddhib/Desktop/Fall2021/STA137/Project1")
Xt = read.csv("Project_1_Data.csv", header = FALSE)
ts.plot(Xt, main="Average Monthly Temperature from January 1964 to December 1975", 
        ylab= "Temperature")
```


## (b) Fit the regression model: seasonal month component plus noise

```{r}
Xt = ts(Xt, start=1964, frequency=12)
Q = factor(cycle(Xt))
reg = lm(Xt ~ 0  + Q, na.action = NULL)
summary(reg)
```
(i)  If the model is correct, determine if the average temperature increase or decrease from month to month? Determine this from January to December.
```{r}
sumDiffs = 0
for (i in (1:length(reg$coefficients)-1)){
        toadd = (reg$coefficients[i] - reg$coefficients[i+1])
        #print(toadd)
        sumDiffs = sumDiffs + toadd
}
avg = sumDiffs/length(reg$coefficients)

```

(ii)
```{r}
plot.ts(cbind(Xt,reg$fitted.values),plot.type = 'single',
        col=c('red',1),
        ylab='Temperature',
        lwd=c(1.5,1.5)) 
legend('bottomright',legend = c('Xt','regression by month'),
       col=c('red',1),
       lty=c(1,1), lwd=c(2,2))
```

```{r}
par(mfrow=c(1,2))
plot(Xt-fitted(reg), main="plot of residuals")
acf(Xt-fitted(reg))
```
The model appears to fit the data well because the residuals look white centered around 0. The acf of the residuals also looks like that of a white noise process because the acf of lag not equal to 0 is within the confidence levels, there fore not significant. The acf at lag 0 is 1 as expected in a white noise process.

## c
```{r}
t = 1:length(Xt)
B11 = cos(2*pi*t/12)
B21 = cos(4*pi*t/12)
B31 = cos(6*pi*t/12)
B41 = cos(8*pi*t/12)
B51 = cos(10*pi*t/12)
B61 = cos(12*pi*t/12)

B12 = sin(2*pi*t/12)
B22 = sin(4*pi*t/12)
B32 = sin(6*pi*t/12)
B42 = sin(8*pi*t/12)
B52 = sin(10*pi*t/12)
B62 = sin(12*pi*t/12)

preg = lm(Xt ~ B11+ B21 + B31 + B41 + B51 + B61 + B12+ B22 + B32 + B42 + B52 + B62)
summary(preg)

```
There are significant frequencies(at a significance of 0.1) are the following coefficients: $\beta_{11}$ , $\beta_{41}$, $\beta_{12}$, $\beta_{42}$ The intercepts of the most significance are coefficients of frequencies with a period length of one year. Therefore, it takes a year for the time series to repeat itself.

(ii)
```{r}
plot.ts(cbind(Xt,preg$fitted.values),plot.type = 'single',
        col=c('red',1),
        ylab='Temperature',
        lwd=c(1.5,1.5)) 
legend('bottomright',legend = c('Xt','Period cycle model'),
       col=c('red',1),
       lty=c(1,1), lwd=c(2,2))
```

(iii)

```{r}
par(mfrow=c(1,2))
plot(Xt-fitted(preg), main="plot of residuals")
acf(Xt-fitted(preg))
```
The model appears to fit the data well because the residuals look white centered around 0. The acf of the residuals also looks like that of a white noise process because the acf of lag not equal to 0 is within the confidence levels, there fore not significant. The acf at lag 0 is 1 as expected in a white noise process.







